The influence of habitat quality and environmental transmission on host-pathogen metapopulation dynamics

### 2014-10-15 Manuscript/manuscript.tex
Edits and figure updates
Some edits from Colleen and changes to make figures more clear.

\title{Environmental pathogen reservoirs and habitat heterogeneity in a metapopulation}

@@

\titlerunning{Pathogen Reservoirs}

@@

Many, if not all, of the pathogens listed above affect spatially structured host populations (e.g. plague in prairie dog colonies, \cite{George2013}), and thus understanding how environmental transmission interacts with patterns of habitat quality is critical to managing disease in these systems.  In this study then, we seek to explore how a pathogen's environmental longevity (how long it can persist and remain infections in the environment), and the distribution of habitat quality in a metapopulation interact to influence pathogen spread and host occupancy, with a specific interest in the roles played by high and low quality habitat.

@@

To address the above questions, we developed a theoretical stochastic patch occupancy model (SPOM) in which each patch can be in one of three possible states: occupied by susceptible hosts (S), occupied by infectious hosts (I), and unoccupied by the host ($\emptyset$).  State transitions are governed by host colonization, extinction, and infection rates, where a susceptible population can become infectious either through direct contact with infectious immigrants, or through a local pathogen reservoir (Table~\ref{transitions}).

@@

The extinction rate of patch $i$ in state $X$ is given by:
\begin{equation}
E_{Xi}=\frac{e_X}{A_i^\alpha},
\end{equation}
where $e_X$ is the extinction rate of a patch of unit quality in state $X$ (here we assume that $e_I>e_S$ to account for disease-induced mortality), and $\alpha$ controls how extinction rate scales with patch quality.  When an infectious population goes extinct, we assume that the hosts leave behind an infectious pathogen reservoir on the patch.

@@

\begin{table}[h!]   
\caption{Parameters of SPOM model, their meaning, and the values assigned under different parameterizations.  Empty cells indicate same value as the default parameterization.}
\begin{tabular}{l p{5cm} rrrr}
Parameter & Interpretation &  Default & Lattice & Equal Ext. & Weak Direct Inf.\\
\hline
$\xi_{im}$ & Scaling parameter for effect of target patch quality on immigration & 0.5 & -- & -- & --\\
$\xi_{em}$ & Scaling parameter for effect of target patch quality in emigration & 0.5 & -- & -- & --\\
$D$& Inverse of mean dispersal distance & 5 & 2 & -- & --\\
$d_{ij}$ & Distance between patch $i$ and $j$ & 1 $\forall i \neq j$ & 1 $\forall i, j$ neighbors & -- & --\\
\hline
$e_S$ & Extinction rate of unit quality susceptible patch & 0.1 & -- & -- & --\\
$e_I$ & Extinction rate of unity quality infectious patch & 0.5 & -- & -- & --\\
$\alpha$ & Strength of environmental stochasticity & 1 & -- & 0 & --\\
\hline
$\delta$ & Probability of direct infection & 0.5 & -- & -- & 0.1\\
$\gamma_0$ & Initial rate of infection from reservoir patch & 0.5 & -- & -- & --
\end{tabular}
\label{params}
\end{table}

@@

The baseline model above assumes that all patches are equally accessible from all other patches (i.e. are separated by distance 1), so to examine the effects of a more rigid spatial structure, we also implemented a model with patches arranged in a square lattice such that each patch is only accessible from its four neighboring patches (Lattice, Table~\ref{params}).  In addition, we explored a model with $\alpha = 0$ so that the extinction rate was constant with patch quality, simulating the effects of high environmental stochasticity, i.e. environmental stochasticity is strong enough that the population size of a patch does not affects its extinction rate (Equal extinction, Table~\ref{params}).  Lastly, we parameterized a model with $\delta = 0.1$ to reduce the relative importance of direct transmission and explore the disease dynamics when transmission is driven largely by the environmental reservoir (Weak direct infection, Table~\ref{params}).

@@

For each of the above parameterizations, we performed two different simulation experiments to explore the combined effects of pathogen environmental longevity and habitat quality distribution on disease dynamics and metapopulation occupancy.  In the first of these, we investigated the effect of increasing habitat heterogeneity on the dynamics of pathogens with a range of environmental longevities.  For these simulations, we explored 10 values of pathogen longevity ranging from 20 to 200 time steps, with values representing the half-life of the pathogen's infectivity, and 10 values for the variance of the patch quality distribution, ranging from 0.02 to 0.2.  For each simulation (100 replicates for each of the 100 longevity-variance combinations), qualities for 100 patches were drawn from a uniform distribution with fixed mean 1 and given variance. In each case, an entirely susceptible population was simulated until it reached an approximate steady state, at which point a randomly chosen occupied patch was infected.  The state of the metapopulation was then tracked for 5000 time steps.  We recorded the proportion of patches in each state at the end of each simulation and classified the simulation as host extinction, disease-free (no infectious populations after 5000 time steps), endemic (both susceptible and infectious populations present), or pandemic (only infectious populations).  In addition, to identify the roles played by individual patches of varying quality, for every simulation, we recorded the quality of each patch, the proportion of time spent in each state (i.e. occupied by susceptible hosts, occupied by infectious hosts, and unoccupied), and the number of transitions between each state.

The above simulations held the mean of patch quality distribution constant and explored the effect of increasing habitat heterogeneity around that mean (i.e. by expanding the quality distribution symmetrically about 1).  To further parse out the relative influence of high and low quality patches separately, we performed a second series of simulations where we expanded the range of the uniform quality distribution to skew towards either high or low quality patches (i.e. by expanding the quality distribution asymmetrically about 1).  For these simulations, at each of the 10 longevities used above, we simulated 100 metapopulations for each of four patch quality distributions: low variance, (patch qualities ranging 0.75 to 1.25); intermediate variance, high quality (range 0.75 to 1.75); intermediate variance, low quality (range 0.25 to 1.25); and high variance (range 0.25 to 1.75).

@@

Since a patch's connectivity is determined by its quality (Eqn~\ref{connectivity}), high quality patches function as highly connected hubs in the metapopulation.  Specifically, high quality patches both attract more colonists from other patches and produce more colonists that spread to other patches.  These two properties have important implications for pathogen spread in the metapopulation. Attracting colonists helps position high quality patches as the ecological "traps" hypothesized above, wherein susceptible hosts repeatedly colonize high quality patches and subsequently become infected (Fig~\ref{infections}).  This infection pressure on high quality patches results from both transmission from a persistent environmental reservoir and from direct transmission from infectious immigrants attracted to the high quality patch, reflecting results from the contact network literature that show that highly connected nodes have higher infection risk (\cite{Christley2005}, \cite{Keeling2005}).

@@

In contrast to high quality habitat, low quality patches help to limit pathogen spread and increase susceptible occupancy (Fig~\ref{sens}). Even though individual low quality patches, due to their high extinction rates and low colonization rates, are unable to support consistent occupancy (Fig~\ref{simvis}), the presence of low quality patches increases overall susceptible occupancy relative to a lower variance metapopulation (Fig~\ref{sens}).  This phenomenon represents the other side of the hub role played by high quality patches in that low quality patches are relatively weakly connected to the metapopulation and are therefore infrequently colonized by infectious individuals.  Moreover, the instability of low quality patches (i.e. the high extinction rate), together with the increased mortality of infectious populations, means that infectious populations do not persist long on low quality patches.  Because low quality patches are so infrequently colonized, at low to intermediate longevities, the environmental reservoir left behind by these infectious populations generally decays before it has the opportunity to infect new susceptible colonists.  As a result, low quality patches effectively represent a dead-end for the pathogen, reducing the number of patches from which it can spread, and thereby reducing infectious occupancy (Fig~\ref{sens}).

@@

However, in cases where the pathogen is able to spread more easily (i.e. at an endemic level), the addition of high quality patches functions similarly to increasing movement in the model developed by \cite{Hess1996}: increasing pathogen spread and as a consequence reducing overall occupancy (Fig~\ref{sens}c).  In these cases, where pathogen longevity is intermediate, or dispersal is limited by metapopulation structure (e.g. a lattice), rather than serving as sources in the metapopulation, high quality patches become sinks.  More importantly, they become ecological traps, which can be even more detrimental to the overall population (\cite{Kristan2003}).  From a management perspective, it is critical to understand this shift, as previous theoretical work suggests that management should focus on maintaining patches where conditions are most favorable (\cite{Strasser2010}).  Generally, this means that managers should focus on maintaining high quality habitat, but Strasser et al. (2010) show that stochastic disturbance (e.g. disease-induced mortality) can lead to cases where focusing on low quality patches is more effective in increasing population growth rate.  Indeed, in these situations, our model suggests that low quality patch distributions, by hindering the spread of the pathogen and providing refuges for susceptible hosts, can produce higher occupancy than high quality distributions.

@@

@@

\begin{figure}[h!]
\centering
\includegraphics[height=5.5cm]{metapop(lattice).pdf}
\caption{Mean occupancy of (a) susceptible patches, (b) infectious patches, and (c) both, as a function of the pathogen's environmental longevity (half-life of infectivity decay) and the variance of the patch quality distribution.}
\label{poutcome_lattice}
\end{figure}

\begin{figure}
\centering
\includegraphics[height=7cm]{highlow(lattice).pdf}
\caption{The effect of the presence of high and low quality patches on (a) mean susceptible occupancy, (b) mean infectious occupancy, and (c) mean total occupancy.  Red solid lines show hiqh quality metapopulations, blue dashed show low quality, black shows high variance, and gray shows low variance.}
\label{sens_lattice}
\end{figure}

\begin{figure}
\centering
\includegraphics[height=7cm]{infevents(lattice).pdf}
\caption{The number of infection events ($S$ to $I$ transitions) for individual patches in endemic simulations.  Light gray lines show smoothed fit for single simulations, while the thick black line shows the smoothed trend over all simulations.}
\label{infections_lattice}
\end{figure}

@@

\begin{figure}[h!]
\centering
\includegraphics[height=5.5cm]{metapop(alpha0).pdf}
\caption{Mean occupancy of (a) susceptible patches, (b) infectious patches, and (c) both, as a function of the pathogen's environmental longevity (half-life of infectivity decay) and the variance of the patch quality distribution.}
\label{poutcome_x0}
\end{figure}

\begin{figure}
\centering
\includegraphics[height=7cm]{highlow(alpha0).pdf}
\caption{The effect of the presence of high and low quality patches on (a) mean susceptible occupancy, (b) mean infectious occupancy, and (c) mean total occupancy.  Red solid lines show hiqh quality metapopulations, blue dashed show low quality, black shows high variance, and gray shows low variance.}
\label{sens_x0}
\end{figure}

\begin{figure}
\centering
\includegraphics[height=7cm]{infevents(alpha0).pdf}
\caption{The number of infection events ($S$ to $I$ transitions) for individual patches in endemic simulations.  Light gray lines show smoothed fit for single simulations, while the thick black line shows the smoothed trend over all simulations.}
\label{infections_x0}
\end{figure}

@@

\begin{figure}[h!]
\centering
\includegraphics[height=5.5cm]{metapop(delta).pdf}
\caption{Mean occupancy of (a) susceptible patches, (b) infectious patches, and (c) both, as a function of the pathogen's environmental longevity (half-life of infectivity decay) and the variance of the patch quality distribution.}
\label{poutcome_delta}
\end{figure}

\begin{figure}
\centering
\includegraphics[height=7cm]{highlow(delta).pdf}
\caption{The effect of the presence of high and low quality patches on (a) mean susceptible occupancy, (b) mean infectious occupancy, and (c) mean total occupancy.  Red solid lines show hiqh quality metapopulations, blue dashed show low quality, black shows high variance, and gray shows low variance.}
\label{sens_delta}
\end{figure}

\begin{figure}
\centering
\includegraphics[height=7cm]{infevents(delta).pdf}
\caption{The number of infection events ($S$ to $I$ transitions) for individual patches in endemic simulations.  Light gray lines show smoothed fit for single simulations, while the thick black line shows the smoothed trend over all simulations.}
\label{infections_delta}
\end{figure}

>>>>

\title{The influence of habitat quality and environmental transmission on host-pathogen metapopulation dynamics}

@@

\titlerunning{Habitat quality and environmental transmission}

@@

Many, if not all, of the pathogens listed above affect spatially structured host populations (e.g. plague in prairie dog colonies, \cite{George2013}), and thus understanding how environmental transmission interacts with patterns of habitat quality is critical to managing disease in these systems.  In this study then, we seek to explore how a pathogen's environmental longevity (how long it can persist and remain infectious in the environment), and the distribution of habitat quality in a metapopulation interact to influence pathogen spread and host occupancy, with a specific interest in the roles played by high and low quality habitat.

@@

To address the above questions, we developed a stochastic patch occupancy model (SPOM) in which each patch can be in one of three possible states: occupied by susceptible hosts (S), occupied by infectious hosts (I), and unoccupied by the host ($\emptyset$).  State transitions are governed by host colonization, extinction, and infection rates, where a susceptible population can become infectious either through direct contact with infectious immigrants, or through a local pathogen reservoir (Table~\ref{transitions}).

@@

The extinction rate of patch $i$ in state $X$ is given by:
\begin{equation}
E_{Xi}=\frac{e_X}{A_i^\alpha},
\end{equation}
where $e_X$ is the extinction rate of a patch of unit quality in state $X$ (here we assume that $e_I \geq e_S$ to account for disease-induced mortality), and $\alpha$ controls how extinction rate scales with patch quality.  When an infectious population goes extinct, we assume that the hosts leave behind an infectious pathogen reservoir on the patch.

@@

\begin{table}[h!]   
\caption{Parameters of the SPOM model, their meaning, and the values assigned under different parameterizations.  Empty cells indicate the same value as the default parameterization.}
\begin{tabular}{l p{5cm} rrrr}
Parameter & Interpretation &  Default & Lattice & Equal Ext. & Weak Direct Inf.\\
\hline
$\xi_{im}$ & Scaling parameter for effect of target patch quality on immigration & 0.5 & -- & -- & --\\
$\xi_{em}$ & Scaling parameter for effect of target patch quality on emigration & 0.5 & -- & -- & --\\
$D$& Inverse of mean dispersal distance & 5 & 2 & -- & --\\
$d_{ij}$ & Distance between patch $i$ and $j$ & 1 $\forall i \neq j$ & 1 $\forall i, j$ neighbors & -- & --\\
\hline
$e_S$ & Extinction rate of unit quality susceptible patch & 0.1 & -- & -- & --\\
$e_I$ & Extinction rate of unity quality infectious patch & 0.5 & -- & -- & --\\
$\alpha$ & Strength of environmental stochasticity & 1 & -- & 0 & --\\
\hline
$\delta$ & Probability of direct infection & 0.5 & -- & -- & 0.1\\
$\gamma_0$ & Initial rate of infection from reservoir patch & 0.5 & -- & -- & --
\end{tabular}
\label{params}
\end{table}

@@

The baseline model above assumes that all patches are equally accessible from all other patches (i.e. are separated by distance 1), so to examine the effects of a more rigid spatial structure, we also implemented a model with patches arranged in a square lattice such that each patch is only accessible from its four neighboring patches (Lattice, Table~\ref{params}).  In addition, we explored a model with $\alpha = 0$ so that the extinction rate was constant with patch quality, simulating the effects of high environmental stochasticity, i.e. environmental stochasticity is strong enough that the population size of a patch does not affect its extinction rate (Equal extinction, Table~\ref{params}).  Lastly, we parameterized a model with $\delta = 0.1$ to reduce the relative importance of direct transmission and explore the disease dynamics when transmission is driven largely by the environmental reservoir (Weak direct infection, Table~\ref{params}).

@@

For each of the above parameterizations, we performed two different simulation experiments to explore the combined effects of pathogen longevity and habitat quality on disease dynamics and metapopulation occupancy.  In the first of these, we investigated the effect of increasing habitat heterogeneity on the dynamics of pathogens with a range of environmental longevities.  For these simulations, we explored 10 values of pathogen longevity ranging from 20 to 200 time steps, with values representing the half-life of the pathogen's infectivity, and 10 values for the variance of the patch quality distribution, ranging from 0.02 to 0.2.  For each simulation (100 replicates for each of the 100 longevity-variance combinations), qualities for 100 patches were drawn from a uniform distribution with fixed mean 1 and given variance. In each case, an entirely susceptible population was simulated until it reached an approximate steady state, at which point a randomly chosen occupied patch was infected.  The state of the metapopulation was then tracked for 5000 time steps.  We recorded the proportion of patches in each state at the end of each simulation and classified the simulation as host extinction, disease-free (no infectious populations after 5000 time steps), endemic (both susceptible and infectious populations present), or pandemic (only infectious populations).  In addition, to identify the roles played by individual patches of varying quality, for every simulation, we recorded the quality of each patch, the proportion of time spent in each state (i.e. occupied by susceptible hosts, occupied by infectious hosts, and unoccupied), and the number of transitions between each state.

The above simulations held the mean of the patch quality distribution constant and explored the effect of increasing habitat heterogeneity around that mean (i.e. by expanding the quality distribution symmetrically about 1).  To further parse out the relative influence of high and low quality patches separately, we performed a second series of simulations where we expanded the range of the uniform quality distribution to favor either high or low quality patches (i.e. by expanding the quality distribution asymmetrically about 1).  For these simulations, at each of the 10 longevities used above, we simulated 100 metapopulations for each of four patch quality distributions: low variance, (patch qualities ranging 0.75 to 1.25); intermediate variance, high quality (range 0.75 to 1.75); intermediate variance, low quality (range 0.25 to 1.25); and high variance (range 0.25 to 1.75).

@@

Since a patch's connectivity is determined by its quality (Eqn~\ref{connectivity}), high quality patches function as highly connected hubs in the metapopulation.  Attracting colonists helps position high quality patches as the ecological "traps" hypothesized above, wherein susceptible hosts repeatedly colonize high quality patches and subsequently become infected (Fig~\ref{infections}).  This increased infection pressure results from both transmission from a persistent environmental reservoir and from direct transmission from infectious immigrants attracted to the high quality patch, reflecting results from the contact network literature that show that highly connected nodes have higher infection risk (\cite{Christley2005}, \cite{Keeling2005}).

@@

In contrast to high quality habitat, low quality patches help to limit pathogen spread and increase susceptible occupancy (Fig~\ref{sens}). Even though individual low quality patches, due to their high extinction rates and low colonization rates, are unable to support consistent occupancy (Fig~\ref{simvis}), the presence of low quality patches increases overall susceptible occupancy relative to a lower variance metapopulation (Fig~\ref{sens}).  This phenomenon represents the other side of the hub role played by high quality patches in that low quality patches are relatively weakly connected to the metapopulation and are therefore infrequently colonized by infectious individuals.  Moreover, the instability of low quality patches (i.e. the high extinction rate), together with the increased mortality of infectious populations, means that infectious populations do not persist long on low quality patches.  Because low quality patches are so infrequently colonized, at relatively low to intermediate longevities, the environmental reservoir left behind by these infectious populations generally decays before it has the opportunity to infect new susceptible colonists.  As a result, low quality patches effectively represent a dead-end for the pathogen, reducing the number of patches from which it can spread, and thereby reducing infectious occupancy (Fig~\ref{sens}).

@@

However, in cases where the pathogen is able to spread more easily (i.e. at an endemic level), the addition of high quality patches functions similarly to increasing movement in the model developed by \cite{Hess1996}: increasing pathogen spread and as a consequence reducing overall occupancy (Fig~\ref{sens}c).  In these cases, where pathogen longevity is intermediate, or dispersal is limited by metapopulation structure (e.g. a lattice), rather than serving as sources that increase metapopulation occupancy, high quality patches become sinks.  More importantly, they become ecological traps, which can be even more detrimental to the overall population (\cite{Kristan2003}).  From a management perspective, it is critical to understand this shift, as previous theoretical work suggests that management should focus on maintaining patches where conditions are most favorable (\cite{Strasser2010}).  Generally, this means that managers should focus on maintaining high quality habitat, but Strasser et al. (2010) show that stochastic disturbance (e.g. disease-induced mortality) can lead to cases where focusing on low quality patches is more effective in increasing population growth rate.  Indeed, in these situations, our model suggests that low quality patch distributions, by hindering the spread of the pathogen and providing refuges for susceptible hosts, can produce higher occupancy than high quality distributions.

@@

\section{Acknowledgements}

CBL would like to thank members of the Webb Lab, past and present, for several rounds of useful feedback.  This work was supported by USGS Cooperative Agreement 101485, and NSF Graduate Research Fellowship, No. DGE-1321845.

@@

\begin{figure}[h!]
\centering
\includegraphics[height=5.5cm]{metapop(lattice).pdf}
\caption{Metapopulation with lattice structure.  Mean occupancy of (a) susceptible patches, (b) infectious patches, and (c) both, as a function of the pathogen's environmental longevity (half-life of infectivity decay) and the variance of the patch quality distribution.}
\label{poutcome_lattice}
\end{figure}

\begin{figure}
\centering
\includegraphics[height=7cm]{highlow(lattice).pdf}
\caption{Metapopulation with lattice structure.  The effect of the presence of high and low quality patches on (a) mean susceptible occupancy, (b) mean infectious occupancy, and (c) mean total occupancy.  Red solid lines show hiqh quality metapopulations, blue dashed lines show low quality, black shows high variance, and gray shows low variance.}
\label{sens_lattice}
\end{figure}

\begin{figure}
\centering
\includegraphics[height=7cm]{infevents(lattice).pdf}
\caption{Metapopulation with lattice structure.  The number of infection events ($S$ to $I$ transitions) for individual patches in endemic simulations.  Light gray lines show the smoothed fit for single simulations, while the thick black line shows the smoothed trend over all simulations.}
\label{infections_lattice}
\end{figure}

@@

\begin{figure}[h!]
\centering
\includegraphics[height=5.5cm]{metapop(alpha0).pdf}
\caption{Extinction rate independent of quality.  Mean occupancy of (a) susceptible patches, (b) infectious patches, and (c) both, as a function of the pathogen's environmental longevity (half-life of infectivity decay) and the variance of the patch quality distribution.}
\label{poutcome_x0}
\end{figure}

\begin{figure}
\centering
\includegraphics[height=7cm]{highlow(alpha0).pdf}
\caption{Extinction rate independent of quality.  The effect of the presence of high and low quality patches on (a) mean susceptible occupancy, (b) mean infectious occupancy, and (c) mean total occupancy.  Red solid lines show hiqh quality metapopulations, blue dashed lines show low quality, black shows high variance, and gray shows low variance.}
\label{sens_x0}
\end{figure}

\begin{figure}
\centering
\includegraphics[height=7cm]{infevents(alpha0).pdf}
\caption{Extinction rate independent of quality.  The number of infection events ($S$ to $I$ transitions) for individual patches in endemic simulations.  Light gray lines show the smoothed fit for single simulations, while the thick black line shows the smoothed trend over all simulations.}
\label{infections_x0}
\end{figure}

@@

\begin{figure}[h!]
\centering
\includegraphics[height=5.5cm]{metapop(delta).pdf}
\caption{Weak direct transmission.  Mean occupancy of (a) susceptible patches, (b) infectious patches, and (c) both, as a function of the pathogen's environmental longevity (half-life of infectivity decay) and the variance of the patch quality distribution.}
\label{poutcome_delta}
\end{figure}

\begin{figure}
\centering
\includegraphics[height=7cm]{highlow(delta).pdf}
\caption{Weak direct transmission.  The effect of the presence of high and low quality patches on (a) mean susceptible occupancy, (b) mean infectious occupancy, and (c) mean total occupancy.  Red solid lines show hiqh quality metapopulations, blue dashed lines show low quality, black shows high variance, and gray shows low variance.}
\label{sens_delta}
\end{figure}

\begin{figure}
\centering
\includegraphics[height=7cm]{infevents(delta).pdf}
\caption{Weak direct transmission.  The number of infection events ($S$ to $I$ transitions) for individual patches in endemic simulations.  Light gray lines show the smoothed fit for single simulations, while the thick black line shows the smoothed trend over all simulations.}
\label{infections_delta}
\end{figure}
